Teaching (Digital) Epigraphy

### 2016-02-19 source/48/EAGLE16-lame-et-al-teaching (1).tex
intro updated

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé}\corref{first}
\author[zag]{Ivan Radman-Livaja}
\author[mau]{Bruce Robertson}
\author[isti]{Federico Ponchio}
\address[ilc]{Centre Camille Jullian, MMSH, CNRS and Laboratorio di Cultura Digitale, University of Pisa}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[zag]{Archaeological Museum in Zagreb}
\address[mau]{Mount Allison University}
\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr}

@@

\noindent Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scholarly \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

\noindent The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
dredging of the Kupa river before WWI. Since the dredging was localized in the very centre of the town, i.e. in front
of the old Roman port quarter, it would seem that all the tags come from a limited area. All of those tags are small
lead tablets, of a more or less rectangular shape, pierced with a hole so that the tag could be attached to the bags
containing the merchandise or to the merchandise itself with a small rope or a metal wire. They all carry an
inscription, most of the time on both sides. Those inscriptions are always, up to now, written in capital letters or
the older Roman cursive, sometimes even in a mixture of both. Most of the tags were reused several times and thus they
carry traces of older inscriptions (palimpsests). Those inscriptions usually follow the same model: one side of the tag
carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
abbreviated form, as well as a price - normally expressed in \emph{denarii} or fractions of the \emph{denarius} - and quite often an
indication of quantity or weight.

@@

\begin{figure}
\centering

\end{figure}
Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like \emph{L, LA, LAN, PAN, T, SAG, PAENV, PAL, LO, LOD, LODI, BANA, AB} is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citet{radman_livaja_segestica_2007}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_1984}; \citet[331-334]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.

@@

\noindent In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g.: Fig.~\ref{fig:12582}, no. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{\ The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

\noindent Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
transcription and drawings) and to classify the written signs according to the following chart (see: Fig.~\ref{fig:table1}). Such preliminary
classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.

@@

\noindent From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
great pleasure and pride in their work. One student, whose cumulative grade at university to this point was below
average, excelled in this project and said that this was what he imagined university studies were going to be like. Two
others prepared a presentation based on this effort, and it was accepted as a 15 minute lecture at the university’s
annual undergraduate research seminar. Three students are currently working on an exercise of online edition and publication (\emph{Épigraphie en Réseau} - ISSN: 2108-7156) of a lead tag using the Markout file created (SVG).

@@

\noindent Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
constrained facilities and personnel. In addition, the epigraphic material may be fragile and delicate: the less these
are manipulated by hands the best it is for their preservation. We believe that a process similar to the one described
here will allow us and others to train far more, wherever they may be in the world, while well preserving the
materials. It will represent a useful intermediary step that develops both epigraphic and digital skills. However,
adding such a step will require us to renew, in some regards, the way we train learners and to reconsider who teaches
what and how. The benefits of this renewal, though, would be many: it would ensure the transmission of epigraphic
tradition and the acquisition of good digital skills, whether to train eminent specialists or to allow access to the
roots of societies to the greatest number.

\vspace{10pt}
\section*{Aknowledgment}
\noindent We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.

@@

>>>>

%%%%%%%%
%TITLE:%
%%%%%%%%
\title{Teaching (Digital) Epigraphy}
\author[ilc]{Marion Lamé\corref{first}}
\author[isti]{Federico Ponchio}
\author[mau]{Bruce Robertson}
\author[zag]{Ivan Radman-Livaja}
\address[ilc]{Centre Camille Jullian, MMSH, CNRS and Laboratorio di Cultura Digitale, University of Pisa}
\address[isti]{Visual Computing Laboratory, ISTI-CNR}
\address[mau]{Mount Allison University}
\address[zag]{Archaeological Museum in Zagreb}
\cortext[first]{Corresponding author. Email: mlame@mmsh.univ-aix.fr. The poster pertaining to this article can be viewed in the \emph{poster} section of the academic blog \emph{Épigraphie En Réseau} -  \url{http://eer.hypotheses.org}.}

@@

Although undergraduate students are naturally attracted to inscriptions and epigraphy affords them an important window
on the past, they are seldom given the opportunity to study inscriptions directly. This is because epigraphical work is
seen as the specialist’s domain: the analysis of such material generally requires a rather high level of expertise,
normally acquired during graduate studies and beyond. Lacking sufficient skills and knowledge to comprehensively
understand often complex epigraphic data, undergraduate students are simply unable to offer the expert opinions
sometimes sought by a project manager. Thus, offering undergraduate students the opportunity to tackle tasks usually
reserved to their senior colleagues is certainly not a common occurrence. In our small project ``Pedagogical,
Scholarly \& Technical Experiment in Digital Epigraphy: the Study Case of the \emph{Tesserarum Sisciae Sylloge} (TSS) through
a Digital Autoptic Process (DAP)'' we aimed to challenge this state of affairs. Using digital methods and a
custom-designed course of rapid study, we offered undergraduates the opportunity to tackle epigraphic tasks usually
reserved to their senior colleagues. We made use of newly-digitized epigraphic material: Roman commercial lead tags
from the ancient city of Siscia.

@@

The Roman Department of the Archaeological Museum in Zagreb (AMZ) contains close to 1200 of these inscribed lead tags
found in Sisak (Siscia), one of the largest urban centers in south-western Pannonia. Most of them were found during the
dredging of the Kupa river before WWI. Since the dredging was localized in the very centre of the town, i.e. in front
of the old Roman port quarter, it would seem that all the tags come from a limited area. All of those tags are small
lead tablets, of a more or less rectangular shape, pierced with a hole so that the tag could be attached to the bags
containing the merchandise or to the merchandise itself with a small rope or a metal wire. They all carry an
inscription, most of the time on both sides. Those inscriptions are always, up to now, written in capital letters or
the older Roman cursive, sometimes even in a mixture of both. Most of the tags were reused several times and thus they
carry traces of older inscriptions (palimpsests). Those inscriptions usually follow the same model: one side of the tag
carries personal names, the other side carries an inscription mentioning the merchandise, most of the time in an
abbreviated form, as well as a price - normally expressed in \emph{denarii} or fractions of the \emph{denarius} - and quite often an
indication of quantity or weight.

@@

Most of the tags, if not all, are linked to the wool trade and the textile industry. Words like \emph{LANA, PAN(N)UM, TVNICA, SAGVM, P(A)ENVLA, PAL(L)A, PALLIOLUM, LODIX, BANATA}, and \emph{ABOLLA}, appear regularly enough without being abbreviated and
thus the interpretation of common abbreviations like \emph{L, LA, LAN, PAN, T, SAG, PAENV, PAL, LO, LOD, LODI, BANA, AB} is
not in doubt. The other abbreviations are mostly related to terms of colour. The prices on those tags are an important
information: they indicated the value of the goods or the cost of a given service (e.g. cleaning, fulling, dyeing). It
would also appear that they were used by fullers and dyers as ownership tags. By noting the name of the client as well
as the type of cloth or service and the price on the tag which was subsequently attached to the item to be processed
(i.e. cleaned or dyed), shop owners could easily return their property to the clients as well as charge them the
correct fee\footnote{\citet{radman_livaja_segestica_2007}, \citet{radman-livaja_roetiquettes_2010}, \citet{radman-livaja_5_2013}, \citet[165-172]{radman-livaja_two_2013}, \citet{radman-livaja_plombs_2014}, for analogous tags see:\citet[97-104]{mocsy_olom_1956}; \citet[195-210]{egger_funf_1968}; \citet[127-138]{frei-stolba_les_1984}; \citet[331-334]{frei-stolba_les_2011}; \citet[121-137]{schwinden_romerzeitliche_1985}; \citet[215-222]{schwinden_zwei_1993};\citet{romer-martijnse_romerzeitliche_1990}; \citet[5-48]{romer-martijnse_fruhkaiserzeitliche_1997}; \citet[301-305]{feugere_etiquette_1993}; \citet[211-220]{weiss_bleietiketten_1991}; \citet[29-40]{paci_etichette_1995}; \citet[207-216]{bassi_tre_1996}; \citet[121-135]{bizzarini_quattro_2005}; \citet[43-51]{buchi_etichette_2005}; \citet[42-110]{cresci_supellex_2010}; \citet[295-317]{jacques_artisanat_2010}; \citet[237-246]{wedenig_bleietikett_2013}.}.

@@

In fact, the transcription and interpretation of Roman inscribed lead tags are often challenging, and so we ignored how
far we could go with these students if useable results would ensue. First we made a selection of fourty tags to be
subjected to the adequate digitization process we chose to work with: Reflectance Transforming Imaging (RTI). It is a
small but nevertheless representative sample of the whole corpus. The students had at their disposal different kinds of
inscriptions, including both easily readable specimens and more challenging graffiti (e.g.: Fig.~\ref{fig:12582}, no. 12582). One may also
profit from the assistance of undergraduate and graduate students, which could not often be the case
before\footnote{The Laboratorio di Cultura Digitale, lead by Prof. Salvatori, at the University of Pisa, applies the
didactic model DIGICRAFT. This means that undergraduate students might share their skills, especially the digital one,
in order to allow an entire team to reach its goal. For more information see: \url{http://www.labcd.unipi.it/laboratorio}.}.
This experiment showed that their contribution can be useful, because technology may compensate for their lack of
knowledge although the final conclusions have to remain in the domain of scholars familiar with the subject. In any
case, this experiment shows how the digital edition may allow both scholars and students to have an open access to what
is basically a primary source of information for humanities (presented as a digital facsimile)\footnote{\citet{gorman_opening_2015}
writes about some of the OA issues: ``While digitisation is not a pre-requisite to gaining access to material (...), and
while digital surrogates of cultural heritage objects do not have to be openly shared once created, just as the
sciences are calling for publication of source data as part of the Open Access movement, opening up access to primary
sources in the cultural heritage sector and encouraging them to be published in a way which is as accessible as
possible has the potential to change the nature of research outputs in the Humanities and Social Sciences, as well as
the nature of research itself in these areas.''}. Of course, learners could train themselves on the drawing of the tag,
but they would not confront the difficulty of interpreting the numerous ambiguities and the responsibility of deciding
what is on the tag.

@@

Once alone, students were asked to ignore the monetary symbols then to decipher (with or without the help of
transcription and drawings) and to classify the written signs according to the following chart (see: Fig.~\ref{fig:table1}). Such preliminary
classification gives a first idea of the main general shapes one can find on a lead tag. It is not a paleographical
study, which has yet to be done, but it is efficient enough to train students at a first level of deciphering.

@@

From the educational standpoint, this pilot project succeeded tremendously. The Canadian students who undertook this
transcription in place of the usual research essay were, on average, far more motivated than their peers, and took
great pleasure and pride in their work. One student, whose cumulative grade at university to this point was below
average, excelled in this project and said that this was what he imagined university studies were going to be like. Two
others prepared a presentation based on this effort, and it was accepted as a 15 minute lecture at the university’s
annual undergraduate research seminar. Three students are currently working on an exercise of online edition and publication (\emph{Épigraphie en Réseau} - ISSN: 2108-7156) of a lead tag using the Markout file created (SVG).

@@

Gaining expertise with material remains requires time. When a museum trains students, this time is often limited due to
constrained facilities and personnel. In addition, the epigraphic material may be fragile and delicate: the less these
are manipulated by hands the best it is for their preservation. We believe that a process similar to the one described
here will allow us and others to train far more, wherever they may be in the world, while well preserving the
materials. It will represent a useful intermediary step that develops both epigraphic and digital skills. However,
adding such a step will require us to renew, in some regards, the way we train learners and to reconsider who teaches
what and how. The benefits of this renewal, though, would be many: it would ensure the transmission of epigraphic
tradition and the acquisition of good digital skills, whether to train eminent specialists or to allow access to the
roots of societies to the greatest number.

\vspace{10pt}
\section*{Aknowledgment}
We wish to acknowledge the research assistance of T. M. Crandall,
History student at Mount Allison University (Sackville, Canada), and C. Le Borgne, a Roman epigrapher and student of
M. Smith, who is Professor of Paleography at the École Nationale des Chartes and at the École Pratique des Hautes
Études (PSL Research University). His current seminar, for the Academic Year 2015/16, at the EPHE is entitled
``La lettre romaine, entre inscriptions monumentales et écritures communes''. The following
students at Mount Allison University participated in this experiment as part of their course work: D. Atkinson, E.
Beaton, M. Gillard, T. Hall and K. Wallace. We warmly thank M. Vazzana and C. Tanca, students of the Laurea
Specialistica in Informatica Umanistica at the University of Pisa, who prepared part of the great pedagogical material
as well as some useful didactic videos for the Laboratorio di Cultura Digitale. We are grateful to the Deutsches
Archäologisches Institut, and especially to Prof. R. Förtsch and to M. Romanello for offering us the opportunity to
present the DAP to the students of the Digital Classicist Berlin Seminar the 8th of December 2015.

@@

%SOSTITUIRE, PRIMA DI MANDARE, CON LA RIGA \bibliography{../../EAGLE}
